---
title: "CW4"
author: "Wen Hans Tan and Eirshad Fahim"
date: "2025-02-30"
output:
  pdf_document: default
  html_document: default
  word_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem 2: Practical (20 marks)

```{r}
library(readr)
moons_data <- read_csv("C:/Users/tanwe/OneDrive/Documents/Stats_Machine_Learning/CW4/moons.csv")

colnames(moons_data)

head(moons_data)
```

(i) **{10 marks}** Load in \$ \texttt{moons.csv}\$ , and use the native
    \$ \texttt{kmeans}\$ function in $R$ to cluster the data into
    $K= \{ 2,3,4,5\}$ components, according to each of the following
    loss functions :

```{r}
#Loading necessary libraries
library(ggplot2)
library(gridExtra)  

set.seed(420)

#Define K values 
K_values <- c(2,3,4,5)

#Seperate Data
X_1 <- moons_data[,1]

#List to store plots 
plots_list <- list()

#Loooping through each value
for (k in K_values){
  #Apply K-Means Clustering Algorithm
  result <- kmeans(X_1, centers=k, nstart=50)
  
  #Store cluster assignments
  moons_data$cluster <- as.factor(result$cluster)
  
  #Scatter Plot for each K-value 
  plot <- ggplot(moons_data, aes(x= X, y= Y, color=cluster))+
          geom_point(size=0.4) +
          ggtitle(paste("Loss-1 Function and K=",k)) +
          labs(x = "X", y= "Y") +
          theme_minimal() +
          theme(plot.margin = margin(1, 1, 1, 1, "pt"))
  
  #Store plot in the list 
  plots_list[[length(plots_list) + 1]] <- plot
}

grid.arrange(grobs = plots_list, nrow = 2, ncol = 2, widths = c(3,3), height = c(3,3))
```

Now, using the Loss-2 Function:

```{r}
#Seperate Data
X_2 <- moons_data[,2]

#List to store plots 
plots_list <- list()

#Define K values 
K_values <- c(2,3,4,5)

#Loooping through each value
for (k in K_values){
  #Apply K-Means Clustering algorithm
  result <- kmeans(X_2, centers=k, nstart=50)
  
  #Store cluster assignments
  moons_data$cluster <- as.factor(result$cluster)
  
  #Scatter Plot for each K-value 
  plot <- ggplot(moons_data, aes(x= X, y= Y, color=cluster))+
          geom_point(size=0.4) +
          ggtitle(paste("Loss-2 Function and K=",k)) +
          labs(x = "X", y= "Y") +
          theme_minimal() +
          theme(plot.margin = margin(1, 1, 1, 1, "pt"))
  
  #Store plot in the list 
  plots_list[[length(plots_list) + 1]] <- plot
}

grid.arrange(grobs = plots_list, nrow = 2, ncol = 2, widths = c(3,3), height = c(3,3))
```

Now, using the Loss-3 Function, where the full feature vector is
utilised:

```{r}
#List to store plots 
plots_list <- list()

#Define K values 
K_values <- c(2,3,4,5)

#Loooping through each value
for (k in K_values){
  #Apply K-Means Clustering Algorithm
  result <- kmeans(moons_data, centers=k, nstart=50)
  
  #Store cluster assignments
  moons_data$cluster <- as.factor(result$cluster)
  
  #Scatter Plot for each K-value 
  plot <- ggplot(moons_data, aes(x= X, y= Y, color=cluster))+
          geom_point(size=0.4) +
          ggtitle(paste("Loss-3 Function and K=",k)) +
          labs(x = "X", y= "Y") +
          theme_minimal() +
          theme(plot.margin = margin(1, 1, 1, 1, "pt"))
  
  #Store plot in the list 
  plots_list[[length(plots_list) + 1]] <- plot
}

grid.arrange(grobs = plots_list, nrow = 2, ncol = 2, widths = c(3,3), heights = c(3,3))
```

Now, we use the last loss function, where there is a transformation:

```{r}
#List to store plots 
plots_list <- list()

#Define K values 
K_values <- c(2,3,4,5)

#Transform Function
transform_to_g <- function(x1,x2) {
  g_x1 <- sqrt((x1 + 1/2)^2 + x2^2)
  g_x2 <- sqrt((x1 - 1/2)^2 + x2^2)
  return(c(g_x1, g_x2))
}

# Apply the transformation using sapply()
transformed_data <- as.data.frame(t(sapply(1:nrow(moons_data), 
                                           function(i) transform_to_g(moons_data[i,1], moons_data[i,2]))))

#Set new column names 
colnames(transformed_data) <- c("gx1", "gx2")

#Convert into numeric data
transformed_data$gx1 <- as.numeric(transformed_data$gx1)
transformed_data$gx2 <- as.numeric(transformed_data$gx2)

#Loooping through each value
for (k in K_values){
  #Apply K-Means Clustering Algorithm
  result <- kmeans(transformed_data, centers=k, nstart=50)
  
  #Store cluster assignments
  transformed_data$cluster <- as.factor(result$cluster)
  
  #Scatter Plot for each K-value 
  plot <- ggplot(transformed_data, aes(x= gx1, y= gx2, color=cluster))+
          geom_point(size=0.4) +
          ggtitle(paste("Loss-4 Function and K=",k)) +
          labs(x = "g(x1)", y= "g(x2)") +
          theme_minimal() +
          theme(plot.margin = margin(1, 1, 1, 1, "pt"))
  
  #Store plot in the list 
  plots_list[[length(plots_list) + 1]] <- plot
}

grid.arrange(grobs = plots_list, nrow = 2, ncol = 2, widths = c(3,3), heights = c(3,3))
```

(ii) **( 5 marks )** Write a function $\verb|cluster_eval|$ which, for a
     given data set $X$ and a range of natural numbers
     $K_{1} \leq K_{2}$ ,
     -   Computes the $K$-Means clusterings for each
         $K_{1} \leq K \leq K_{2}$ , and
     -   Given this collection of clusterings, plots the values of the
         $K$-Means loss function at the optimal clusterings, as a
         function of $K$.

Use your implementation of $\verb|cluster_eval|$ to compare the
clusterings which you obtained in Part (i) of this problem. Do you find
that you different clusterings agree on what the 'correct' number of
clusters in this data are? **):**

```{r}
library(ggplot2)

cluster_eval <- function(K1, K2){
  #Create empty list to store plots 
  plots_list <- list()
  
  # Create empty numeric vectors to store WCSS
  wcss_values <- list(
    wcss_value_1 = numeric(length = K2 - K1 + 1),
    wcss_value_2 = numeric(length = K2 - K1 + 1),
    wcss_value_3 = numeric(length = K2 - K1 + 1),
    wcss_value_4 = numeric(length = K2 - K1 + 1)
  )
  
  #Loop values from K1 to K2
  for (k in K1:K2){
    
    #Apply k-means algorithm to all clusterings
    result_loss_1 <-kmeans(X_1, centers=k, nstart = 50)
    result_loss_2 <- kmeans(X_2, centers=k, nstart = 50)
    result_loss_3 <- kmeans(moons_data, centers=k, nstart = 50)
    result_loss_4 <- kmeans(transformed_data, centers=k, nstart = 50)
    
    #Store the WCSS values
    wcss_values$wcss_value_1[k - K1 + 1] <- result_loss_1$tot.withinss
    wcss_values$wcss_value_2[k - K1 + 1] <- result_loss_2$tot.withinss
    wcss_values$wcss_value_3[k - K1 + 1] <- result_loss_3$tot.withinss
    wcss_values$wcss_value_4[k - K1 + 1] <- result_loss_4$tot.withinss
  }
  
  # Create individual plots for each dataset
  datasets <- list("X_1" = wcss_values$wcss_value_1, 
                   "X_2" = wcss_values$wcss_value_2,
                   "moons_data" = wcss_values$wcss_value_3, 
                   "transformed_data" = wcss_values$wcss_value_4)
  
  for (i in seq_along(datasets)) {
    wcss_df <- data.frame(K = K1:K2, WCSS = datasets[[i]])
    
    plot <- ggplot(wcss_df, aes(x = K, y = WCSS)) +
      geom_line(color = "black") +
      geom_point(size = 1.5, color = "red") +
      ggtitle(paste("WCSS for", names(datasets)[i])) +
      labs(x = "K", y = "(WCSS)") +
      theme_minimal()
    
    # Store each plot in the list
    plots_list[[i]] <- plot
  }
  
  #Arrange plots in a 2x2 grid
  grid.arrange(grobs = plots_list, nrow = 2, ncol = 2, widths = c(3, 3), heights = c(3, 3))
}

#Use cluster_eval function
cluster_eval(2, 16)
```

We plotted the WCSS function against k for all types of loss function.
We can examine the elbow point in the plot ; it looks like the optimal
value of $K$ for their respective loss functions are:

-   $Loss^{1}$ : K= 4

-   $Loss^{2}$ : K= 4

-   $Loss^{3}$ : K = 5

-   $Loss^{4}$ : K= 5

From part (i), depending on the loss function, we can inspect
**visually** what the optimal value of K could be. In my opinion, these
are the optimal values of K for each function:

-   $Loss^{1}$ : K=3

-   $Loss^{2}$ : K= 4

-   $Loss^{3}$ : K = 4

-   $Loss^{4}$ : K= 4

Again, these are completely subjective to each person. Since most of the
"correct" cluster number is differs at most by 1, choices of $K$ chosen
using the elbow method is more or less correct.

iii\. **( 5 marks )**Load in R’s quakes dataset. This dataset stores
information about 1000 seismic events (i.e. earthquakes and similar)
which have occurred in the vicinity of the island of Fiji over the past
60 or so years. The first three columns (lat, long, depth) record the
geographical location of the event, the fourth (mag) records the
strength of the event (as measured on the Richter scale), and the the
fifth (station) records how many stations documented the event in
question.

In the context of this question, we are interested in whether the
locations at which these seismic events take place are spatially
localised, i.e. if they cluster around specific areas of the island.

-   Run a standard K-Means clustering on the full data set for $K =
    \{2, 3, 4, 5 \}$, and plot the outcome, focusing on the
    $\texttt{lat}$ and $\texttt{long}$ dimensions.

-   Does the clustering which you obtain in this way look spatially
    reasonable? Propose an alternative approach to obtaining a
    clustering which is more coherent (still using K-Means).

```{r}
data("quakes")
head(quakes)

#List to store plots 
plots_list <- list()

#Define K values 
K_values <- c(2,3,4,5)

#Loooping through each value
for (k in K_values){
  #Apply K-Means Clustering Algorithm
  result <- kmeans(quakes, centers=k, nstart=50)
  
  #Store cluster assignments
  quakes$cluster <- as.factor(result$cluster)
  
  #Scatter Plot for each K-value 
  plot <- ggplot(quakes, aes(x= lat, y= long, color=cluster))+
          geom_point(size=0.4) +
          ggtitle(paste("Full Data K-Means & K=",k)) +
          labs(x = "Latitude", y= "Longitude") +
          theme_minimal() +
          theme(plot.margin = margin(1, 1, 1, 1, "pt"))
  
  #Store plot in the list 
  plots_list[[length(plots_list) + 1]] <- plot
}

grid.arrange(grobs = plots_list, nrow = 2, ncol = 2, widths = c(3,3), heights = c(3,3))
```

For $K=2$, the clusters are well seperated but there are some
overlapping points ; overall, it looks spatially reasonable. As the
number of clusters increase, they appear more fragmented and
artificially split, especially for $K \in \{4,5 \}$ . An alternative
approach while still using K-Means will be to do k-means with only
$\texttt{lat}$ and $\texttt{long}$ , as including the other features
does not help in finding whether location of the seismic events are
*spatially localised.*

```{r}
#List to store plots 
plots_list <- list()

#Chooses lat and long only
quakes_clean <- subset(quakes, select = -c(depth, mag, stations))

#Loooping through each value
for (k in K_values){
  #Apply K-Means Clustering Algorithm
  result <- kmeans(quakes_clean, centers=k, nstart=50)
  
  #Store cluster assignments
  quakes_clean$cluster <- as.factor(result$cluster)
  
  #Scatter Plot for each K-value 
  plot <- ggplot(quakes_clean, aes(x= lat, y= long, color=cluster))+
          geom_point(size=0.4) +
          ggtitle(paste("Full Data K-Means & K=",k)) +
          labs(x = "Latitude", y= "Longitude") +
          theme_minimal() +
          theme(plot.margin = margin(1, 1, 1, 1, "pt"))
  
  #Store plot in the list 
  plots_list[[length(plots_list) + 1]] <- plot
}

grid.arrange(grobs = plots_list, nrow = 2, ncol = 2, widths = c(3,3), heights = c(3,3))
```

Now, it looks spatially localised! Clusters are more identifiable!
